---
title: ANLYTC3 Mini-Project
Analytics Practitioner: "Zarex Alvin Daria"
output: html_document
---


### Instructions 

The data set `nlschools` (in the `MASS` library, use `data(nlschools, package = "MASS")` to load in the data) contains the following records for 2287 students in the Netherlands (see `?nlschools` for more information) 

* `lang`: their test score on a language exam
* `IQ`: their verbal IQ
* `class`: the ID number for their classroom
* `GS`: the number of students in each class
* `SES`: the socio-economic status of their family
* `COMB`: was the student in a multi-grade class? (0=no, 1=yes)

Your task is to explore and describe this data set. You should use both models and visualization. You should use your judgement as to what are the most important aspects of the data to highlight; however, the following questions are of particular interest:

1. Are there discrepancies in `IQ` or `SES` in the different classes, or when grouping by multi-grade vs non-multi-grade classes?
2. When did students perform better or worse on the language exam? Describe which variables had the most important effects. 
3. Do you think there are interactions in the effects of the variables on the language exam score? Speculate as to the cause of any such effects that you think should be included.

Always think about what you are doing -- blindly replicating the lectures may not result in the best analysis! Every data set is different, so you may need to use your judgement on the right thing to do. Your goal is to understand the data as best as your can, particularly in regards to questions 1-3 above.

NOTE: Since this is an observational study, we'll understand that any claims about causality should be taken mainly as speculation or hypothesis. Usage of language regarding the "effects" of a variable on the outcome of interest will be understood to be informal, so we won't penalize you. (This will probably make the report much easier to write.)

### Presentation Requirements

Your output should be this rmarkdown file, which should compile into a report that gives your analysis. 

The report should be understandable. One possible outline is: 

1. Intro: what are your main steps, and what are your findings?
2. Analysis: Describe each step and why you did it, show the R code and results, then comment on the results.
3. Conclusions: recap and summarize 

Typically, it's best to do step 2 first, and then write step 1 and 3 after you are finished. Bulleted lists may be especially helpful for these parts. 

Code should be clearly commented. All plots should be presentable and properly labeled. Mitigate overplotting whenever possible. When building a model, explain each step so that we know why you are doing it. 

The due date is 11:59pm EST June 28, 2019 (Friday). This is a hard deadline -- no late days may be used. Submit this Rmd file on Canvas Don't submit additional files.

### Report:

```{r echo=FALSE, fig.height=2.2, fig.show="hold", fig.width=4, cache=TRUE, dpi=80}
library("MASS")
library(ggplot2)
library(dplyr, warn.conflicts = FALSE)
library(reshape2)
data <- nlschools

#Count the number of rows in the dataset
total.rows <- count(data)

# Average Language Exam
avg.lang <- mean(data$lang)

# Standard Deviation in the Language Exam
sd.lang <- sd(data$lang)
```

#### Introduction
The dataset that the Analytics Practitioner will be working on is the nlschools dataset which is a study of 2,287 eighth-grade pupils (aged about 11) in 132 classes in 131 schools in the Netherlands conducted by Snijders and Bosker (1999).

To get a glimpse of what the dataset is about, the Analytics Practitioner determined the average values of some attributes as detailed below:

1. The dataset has `r total.rows` rows.
2. The average language exam test score of all the students is `r avg.lang`.
3. The standard deviation of the students' language exam is `r sd.lang`.

The below tables show the summary of the nlschools dataset.

```{r echo=FALSE, fig.height=2.2, fig.show="hold", fig.width=4, cache=TRUE, dpi=80}
# Dataset Summary
data.summary <- summary(nlschools)

knitr::kable(data.summary,
             caption = "Table 1: Summary of the Dataset",
             digits = 2)
```



Table 1 shows the summary of the dataset which details the following:

1. The highest language exam score of a student (lang column) is 58, while its average is 40.93, and the lowest is 9.
2. The highest student IQ (IQ column) is 18, while its average is 11.83, and the lowest is 4.
3. The highest student count (GS column) in a class is 39, while its average is 26.51, and the lowest is 10.
4. The highest socio-economic status (SES column) of a student is 50, while the average is 27.81, and the lowest is 10.
5. There are a total of 629 students who are under a multi-grade class (COMB column), while there are 1,658 students who are not.


#### Body
Now, to expound on the summary of the dataset, the Analytics Practitioner will answer the questions posed using tables and graphs.


1. Are there discrepancies in `IQ` or `SES` in the different classes, or when grouping by multi-grade vs non-multi-grade classes?

```{r echo=FALSE, fig.height=2.2, fig.show="hold", fig.width=4, cache=TRUE, dpi=80}
# Suppress summarise info
options(dplyr.summarise.inform = FALSE)
# Discrepancies by Class
class.disc <- data %>%
    group_by(class, GS, COMB) %>%
    summarise(avg.IQ = mean(IQ), sd.IQ = sd(IQ), avg.SES = mean(SES), sd.SES = sd(SES), avg.lang = mean(lang), sd.lang = sd(lang))
```

```{r echo=FALSE, fig.height=2.2, fig.show="hold", fig.width=4, cache=TRUE, dpi=80}
# Highest standard deviation in IQ
top.IQ.sd <- head(select(
  as_tibble(class.disc %>% 
    arrange(desc(sd.IQ))),
  sd.IQ, avg.IQ, class, GS, avg.lang, sd.lang , avg.SES, sd.SES), n=5)

knitr::kable(top.IQ.sd,
             caption = "Table 2: The classes with the highest standard deviation in IQ",
             digits = 2)
```

```{r echo=FALSE, fig.height=2.2, fig.show="hold", fig.width=4, cache=TRUE, dpi=80}
# Lowest standard deviation in IQ
btm.IQ.sd <- tail(select(
  as_tibble(class.disc %>% 
  arrange(desc(sd.IQ))),
  sd.IQ, avg.IQ, class, GS, COMB, avg.lang, sd.lang , avg.SES, sd.SES), n=5)

knitr::kable(btm.IQ.sd,
             caption = "Table 3: The classes with the lowest standard deviation in IQ",
             digits = 2)
```

```{r echo=FALSE, fig.height=2.2, fig.show="hold", fig.width=4, cache=TRUE, dpi=80}
# Highest standard deviation in SES
top.SES.sd <- head(select(
  as_tibble(class.disc %>% 
    arrange(desc(sd.SES))),
  sd.SES, avg.SES, class, COMB, GS, avg.lang, sd.lang , avg.IQ, sd.IQ), n=5)

knitr::kable(top.SES.sd,
             caption = "Table 4: The classes with the highest standard deviation in SES",
             digits = 2)
```

```{r echo=FALSE, fig.height=2.2, fig.show="hold", fig.width=4, cache=TRUE, dpi=80}
# Lowest standard deviation in SES
btm.SES.sd <- tail(select(
  as_tibble(class.disc %>% 
  arrange(desc(sd.SES))),
  sd.SES, avg.SES, class, GS, COMB, avg.lang, sd.lang , avg.IQ, sd.IQ), n=5)

knitr::kable(btm.SES.sd,
             caption = "Table 5: The classes with the lowest standard deviation in SES",
             digits = 2)
```


```{r echo=FALSE, fig.height=2.2, fig.show="hold", fig.width=4, cache=TRUE, dpi=80}
# Class with the highest language scores
top.lang.scores <- head(select(
  as_tibble(class.disc %>% 
    arrange(desc(avg.lang))),
  avg.lang, sd.lang, class, GS, COMB, avg.IQ, sd.IQ, avg.SES, sd.SES), n=5)

knitr::kable(top.lang.scores,
             caption = "Table 2: The top 5 classes with the highest language scores",
             digits = 2)
```

```{r echo=FALSE, fig.height=2.2, fig.show="hold", fig.width=4, cache=TRUE, dpi=80}
# Class with the highest language scores
bottom.lang.scores <- tail(select(
  as_tibble(class.disc %>% 
    arrange(desc(avg.lang))),
  avg.lang, sd.lang, class, GS, COMB, avg.IQ, sd.IQ, avg.SES, sd.SES), n=5)

knitr::kable(bottom.lang.scores,
             caption = "Table 3: The bottom classes with the lowest language scores",
             digits = 2)
```


Table 2 shows the top 5 classes with the highest number 



```{r echo=FALSE, fig.height=2.2, fig.show="hold", fig.width=4, cache=TRUE, dpi=80}
# Class with the lowest number of students
tail(as_tibble(class.disc %>%
    arrange(desc(GS))), n=5)
```

```{r echo=FALSE, fig.height=2.2, fig.show="hold", fig.width=4, cache=TRUE, dpi=80}
# Class with the highest IQ Average
print(as_tibble(class.disc %>%
    arrange(desc(avg.IQ))), n=5)
```

```{r echo=FALSE, fig.height=2.2, fig.show="hold", fig.width=4, cache=TRUE, dpi=80}
# Class with the lowest IQ Average
tail(as_tibble(class.disc %>%
    arrange(desc(avg.IQ))), n=5)
```

```{r echo=FALSE, fig.height=2.2, fig.show="hold", fig.width=4, cache=TRUE, dpi=80}
# Class with the highest SES Average
print(as_tibble(class.disc %>%
    arrange(desc(avg.SES))), n=5)
```

```{r echo=FALSE, fig.height=2.2, fig.show="hold", fig.width=4, cache=TRUE, dpi=80}
# Class with the lowest SES Average
tail(as_tibble(class.disc %>%
    arrange(desc(avg.SES))), n=5)
```

```{r echo=FALSE, fig.height=2.2, fig.show="hold", fig.width=4, cache=TRUE, dpi=80}
# Class with the highest Language Score Average
print(as_tibble(class.disc %>%
    arrange(desc(avg.lang))), n=5)
```

```{r echo=FALSE, fig.height=2.2, fig.show="hold", fig.width=4, cache=TRUE, dpi=80}
# Class with the lowest Language Score Average
tail(as_tibble(class.disc %>%
    arrange(desc(avg.lang))), n=5)
```

```{r echo=FALSE, fig.height=2.2, fig.show="hold", fig.width=4, cache=TRUE, dpi=80}
# Class name and Average IQ
# class.iq <- ggplot(aes(x = class, y = IQ), data = nlschools)
q <- c(.25, .5, .75, 1)

ggplot(nlschools, aes(x = lang)) +
  geom_histogram(col = "black", binwidth=.5) +
  labs(x = "Language Exam Test Score")

ggplot(nlschools, aes(x = IQ)) +
  geom_histogram(col = "black", binwidth=.5) +
  labs(x = "IQ")

ggplot(nlschools, aes(x = GS)) +
  geom_histogram(col = "black", binwidth=.5) +
  labs(x = "GS")

ggplot(nlschools, aes(x = SES)) +
  geom_histogram(col = "black", binwidth=.5) +
  labs(x = "SES")

```

```{r echo=FALSE, fig.height=2.2, fig.show="hold", fig.width=4, cache=TRUE, dpi=80}
# language score and verbal IQ
#class.ses <- ggplot(aes(x = class, y = SES), data = nlschools)
#class.ses + 
#  stat_summary(fun = "mean", geom = "bar") +
#    labs(
#       x="class",
#       y="SES",
#       title="Average Student SES by Class", 
#       caption="Source: nlschools")

```

2. When did students perform better or worse on the language exam? Describe which variables had the most important effects.

```{r echo=FALSE, fig.height=2.2, fig.show="hold", fig.width=4, cache=TRUE, dpi=80}
data <- nlschools

# language score, verbal IQ by class
p1 <- ggplot(data, mapping = aes(x = lang , y = IQ, colour = IQ))

p1 + 
  geom_point(aes()) +
  geom_smooth(show.legend = TRUE, method = 'gam') +
    labs( x="Language Score",
       y="IQ",
       title="Language Score and IQ", 
       caption="Source: nlschools")

# language score, verbal IQ by COMB
p2 <- ggplot(data, mapping = aes(x = lang , y = GS, colour = GS))
p2 + 
  geom_point(aes()) +
  geom_smooth(show.legend = TRUE, method = 'gam') +
  # stat_qq() +
  # stat_qq_line() +
    labs( x="Language Score",
       y="GS",
       title="Language Score + GS", 
       caption="Source: nlschools")


# language score, SES
p3 <- ggplot(data, mapping = aes(x = lang , y = SES, colour = SES))
p3 + 
  geom_point(aes()) +
  geom_smooth() +
  # stat_qq() +
  # stat_qq_line() +
    labs( x="Language Score",
       y="SES",
       title="Language Score + SES", 
       caption="Source: nlschools")

# language score, verbal IQ by COMB
# p3 <- ggplot(data, mapping = aes(fill=lang, x = lang , y = COMB, colour = COMB))
# p3 + 
#   geom_point(aes()) +
#   geom_smooth() +
#     labs( x="Language Score",
#        y="COMB",
#        title="Language Score + COMB", 
#        caption="Source: nlschools")

#Average language exam test score by COMB
ggplot(aes(fill=lang, x = COMB, y = lang), data = data) + 
  geom_bar(position="dodge", stat="identity")
    
ggplot(aes(fill=lang, x = COMB, y = lang), data = data) + 
  stat_summary(fun = "mean", geom = "bar")

sd.class.lang <- data %>%
    group_by(COMB) %>%
    dplyr::summarise(avg.IQ = mean(IQ), sd.IQ = sd(IQ), avg.SES = mean(SES), sd.ses = sd(SES), avg.lang = mean(lang), sd.lang = sd(lang), count.lang = length(lang))
sd.class.lang  

# sd.class.lang <- data %>%
#     group_by(class, GS) %>%
#     summarise(avg.IQ = mean(IQ), sd.IQ = sd(IQ), avg.SES = mean(SES), sd.ses = sd(SES), avg.lang = mean(lang), sd.lang = sd(lang))

```


3. Do you think there are interactions in the effects of the variables on the language exam score? Speculate as to the cause of any such effects that you think should be included.




#### Conclusion

