<style>
  .reveal pre {
    font-size: 13pt;
  }
  .reveal section p {
    font-size: 32pt;
  }
  .reveal div {
    font-size: 30pt;
  }
  .reveal h3 {
    color: #484848;
    font-weight: 150%;
  }
</style>

Lecture 8: Distributions
====
author: 95-868
font-family: Helvetica
autosize: false
width:1440
height:900


Agenda
=======

* Visualizing a univariate distribution
* Comparing between two groups with QQ plots

Histograms
=======

Histogram of the fuel efficiency for city driving (`cty`) in the `mpg` dataset:

```{r fig.width=6, fig.height=3, dpi=150, cache=FALSE, fig.align="center", message=TRUE}
library(ggplot2) 
library(reshape2)
#load(week4.data.rda)
# geom_histogram makes a histogram from x
ggplot(data=mpg, mapping=aes(x=cty)) + geom_histogram()
```

Histograms (continued)
=====

Note the warning message that we didn't specify a bin width. 

The default bin width in `ggplot2` can me misleading. Various binwidths:

```{r fig.width=4, fig.height=2.2, dpi=80, cache=TRUE, fig.show="hold", echo=FALSE}
# binwidth argument changes the histogram
ggplot(data=mpg, mapping=aes(x=cty)) + geom_histogram(binwidth=.5) + labs(title='binwidth=0.5') 
ggplot(data=mpg, mapping=aes(x=cty)) + geom_histogram(binwidth=1) + labs(title='binwidth=1.0') 
ggplot(data=mpg, mapping=aes(x=cty)) + geom_histogram(binwidth=1.2) + labs(title='binwidth=1.2') 
ggplot(data=mpg, mapping=aes(x=cty)) + geom_histogram(binwidth=1.4) + labs(title='binwidth=1.4') 
ggplot(data=mpg, mapping=aes(x=cty)) + geom_histogram(binwidth=1.6) + labs(title='binwidth=1.6') 
ggplot(data=mpg, mapping=aes(x=cty)) + geom_histogram(binwidth=1.8) + labs(title='binwidth=1.8') 
ggplot(data=mpg, mapping=aes(x=cty)) + geom_histogram(binwidth=2.0) + labs(title='binwidth=2.0') 
ggplot(data=mpg, mapping=aes(x=cty)) + geom_histogram(binwidth=2.2) + labs(title='binwidth=2.2') 
```

* `binwidth=1.0` or `1.2`: bin heights seem to decrease as `cty` increases 
* `binwidth=1.4` or `1.8`: two distinct modes in the distribution
* `binwidth=2.0`: distribution looks flat (i.e., uniform)

Histograms (continued)
======

Hard to say which is bin width true. 

Use caution when interpreting the "shape" of a histogram.

Kernel density estimates have similar weaknesses.

(clarification: they're still quite useful -- just know that care may be needed!)

Quantiles
========

Here are 4 quantiles in `mpg$cty`:

```{r fig.width=4, fig.height=2.5, dpi=70, fig.show = 'hold', cache=TRUE}
# find the 10%, 30%, 50%, and 100% quantiles of mpg$cty:
quantile(mpg$cty, probs = c(.1, .3, .5, 1))
```

So if your city mileage is `r quantile(mpg$cty, probs = .3)`, then you are better than 30% of the cars in the `mpg` dataset.


Quantile Plot
======

**quantile plot**: `y` is the data , `x` is the quantiles. 

```{r fig.width=6, fig.height=3, dpi=150, cache=TRUE, fig.align="center"}
# stat_qq() draws a quantile plot from sample when distribution = "qunif"
ggplot(data=mpg, mapping=aes(sample=cty)) + 
  stat_qq(geom = 'point', distribution = 'qunif') + 
  labs(x='Quantile', y='City MPG', title='Quantile Plot, City MPG')
```


Quantile Plot (continued)
=====

Advantage: more precise than histogram

For example, we know that:

* The median value of `cty` is `17`
* The values of `cty` are not exactly uniform distributed from `11` to `21`, but its fairly close.
* The top 10% values of `cty` are widely dispersed from `21` to `35`. These are the "tails" of the distribution.

Disadvantage: quantile plots are unfamiliar. You may need to explain it to your audience.

Comparing two distributions
=======

Compare birth weights of infants, grouped by smoking status of mother

* `birthwt$bwt`: the birth weight of infant
* `birthwt$smoke`: did the mother smoke during pregnancy?

```{r cache=TRUE, echo = FALSE}
library(MASS) # for birthwt data set
library(plyr) # for mapvalues
# relabel birthwt$smoke so that it is more readable
birthwt$smoke = mapvalues(birthwt$smoke, from = c(0,1), to = c("no", "yes"))
```


Visualizing the quantiles
========

Here are the quantile plots for each group, side-by-side:
```{r fig.width=8, fig.height=3, dpi=150, cache=TRUE, fig.align='center'}
# stat_qq() + facet_wrap() for multiple plots
ggplot(data = birthwt, mapping = aes(sample = bwt)) + 
  stat_qq(geom='point', distribution='qunif') + 
  facet_wrap("smoke") + 
  labs(x='Quantiles', y='Birth Weight (g)', title = 'Quantile Plot, Birth Weight (Grouped by Smoking)')
```

Visualizing the quantiles (continued)
========

Putting them on the same plot is better:

```{r fig.width=6, fig.height=3, dpi=150, cache=TRUE, fig.align='center'}
# stat_qq() with mappings for both sample and color 
ggplot(data = birthwt, mapping = aes(sample = bwt, color=smoke)) + 
  stat_qq(distribution='qunif') + 
  labs(x='Quantiles', y='Birth Weight (g)', color = 'Smoking Status', 
       title = 'Quantile Plot, Birth Weight (Grouped by Smoking)')
```

Difference is small for lowest quantiles (the most important case?)

Quantile-Quantile Plots
======

Problem: differences between curves can be hard to visually measure. 

One solution: 

* Plot quantiles of each group as `x` and `y` in a scatterplot.

Known as a **Quantile-Quantile plot**, or **QQ plot**. 


======

For some, it may help understanding to make a QQ plot "from scratch":

```{r fig.width=6, fig.height=3, dpi=150, cache=TRUE, fig.align='center'}
# how big is the smaller group?
n.pts = with(birthwt, 
             min( length(bwt[ smoke == 'no' ]), length(bwt[ smoke == 'yes']) ) )
# which quantiles should we plot in the QQ plot?
probs = seq(from = 0, to = 1, length.out = n.pts)
# extract the quantiles
q1 = with(birthwt, quantile( bwt[ smoke == 'no'], probs = probs))
q2 = with(birthwt, quantile( bwt[ smoke == 'yes'], probs = probs))
# make a QQ plot
ggplot(mapping = aes(x=q1, y=q2)) + geom_point() + geom_abline(a=0, b=1) + labs(x = 'smoke == no', y ='smoke == yes', title = 'QQ plot, Birth Weight grouped by Smoking Status')
```

=====

Built-in function in base graphics is convenient:

```{r fig.width=4, fig.height=3.5, dpi=150, cache=TRUE, fig.align="center"}
# qqplot() makes QQ plot from x and y vectors
# with() makes it more readable
with(birthwt, 
     qqplot(x = bwt[ smoke == 'no'], y = bwt[ smoke == 'yes'], main='QQ plot'))
# abline() adds a reference line, with y intercept 0 and slope 1
abline(0,1)
```



Reading a QQ plot
======

We can tell which group has bigger quantiles, based on whether the points are above or below the reference line `y=x`. 

We can see in this QQ plot:

* Birth weights are bigger when `smoke == no`
* But the difference goes away at the lower quantiles (i.e. premature births).

Discussion
=====

Advantage of visualization:

You can see not only whether the two samples are different, but in which quantiles they differ specifically. 

The QQ plot is more standard, but the quantile plot might be easier to understand.



Two sample tests
======

For smoking, difference in birth weight is **statistically significant**:

```{r cache=TRUE}
# t-test checks for a difference in means, between x and y vectors, assuming normally distributed data
with(birthwt, 
     t.test(x = bwt[ smoke == "no"], y = bwt[ smoke == "yes"]))
```

See also `wilcox.test` and `ks.test` for non-normal data

Discussion
======

Caveats to the t-test:

* There still might not be any difference at the lowest quantiles 
* It doesn't mean that smoking was necessarily the cause of the difference.

And generally: 
* If the test didn't find anything, it might only be a limitation of the test. 

Tests can tell you if the difference is statistically significant, in some limited way. 

Visualization lets you see what the difference actually is.

QQ plot under additive shift
=======

What if `x` and `y` are related by an additive shift?
```{r fig.width=4, fig.height=3.5, dpi=150, cache=TRUE, fig.align="center"}
# x is standard gaussian, y is standard gaussian + 2
x = rnorm(45)
y = rnorm(45) + 2
# QQplot is approximately y = x + 2 (straight line)
qqplot(x = x, y =y, main='QQplot')
abline(0,1)
```

QQ plot under multiplicative shift
=====
```{r fig.width=3, fig.height=3.5, dpi=150, cache=TRUE, fig.align="center"}
# x is standard gaussian, y is standard gaussian + 2
x = rnorm(80)
y = rnorm(70) * 4
# QQplot is approximately y = x * 4 (also straight line)
qqplot(x = x, y =y, main='QQplot')
abline(0,1)
```


QQ plot may be noisy for small samples
=====


```{r fig.width=3, fig.height=3.5, dpi=150, cache=TRUE, fig.align="center"}
# same as before but less data
x = rnorm(30)
y = rnorm(30) * 4
# should be y = x*4, but line is noisy now
qqplot(x = x, y =y, main='QQplot')
abline(0,1)
```




Theoretical QQ plot
========

Let's compare the quantiles of `birthwt$bwt` to those of a Normal(0,1) distribution, in a QQ plot.

* `quantile(birthwt$bwt, ...)` gives the quantiles of `birthwt$bwt`.
* `qnorm()` returns the quantiles of a normal distribution.

Here's the long way to implement:

```{r fig.width=6, fig.height=4, dpi=150, cache=TRUE, fig.align='center'}
# which quantiles do we want to compare?
n.pts = length(birthwt$bwt)
# we want to compare the quantile corresponding to each data point
probs = seq(from=0, to=1, length.out = n.pts)
# get those quantiles from birthwt$bwt
q1 = quantile(birthwt$bwt, probs = probs)
# get the same quantiles from a normal distribution (new function!)
q2 = qnorm(p = probs)
```

Theoretical QQ plots (continued)
=======

```{r fig.width=6, fig.height=3, dpi=150, cache=TRUE, fig.align='center'}
# make the QQ plot
ggplot(mapping = aes(y=q1, x=q2)) + geom_point() + 
  geom_abline(intercept=2944, slope=729) + labs(x = 'Normal Quantiles', y ='Birthweight', title = 'QQ plot, Birth Weight vs Normal Distribution') + annotate(geom = 'text', x = 0, y=1000, label='Ref. Line: y = 729 x + 2944', hjust=0)
```


=======

`stat_qq(dist = qnorm)` gives the same plot:

```{r fig.width=6, fig.height=3, dpi=150, cache=TRUE, fig.align='center'}
ggplot(data = birthwt, mapping=aes(sample = bwt)) + stat_qq(distribution='qnorm') + 
  geom_abline(intercept=2944, slope=729) + 
  labs(x = 'Normal Quantiles', y ='Birthweight', title = 'QQ plot, Birth Weight vs Normal Distribution') + 
  annotate(geom = 'text', x = 0, y=1000, label='Ref. Line: y = 729 x + 2944', hjust=0)
```

The points are close to the line `y= 729x + 2944`. What does this mean? 

(**hint:** what is the mean and standard deviation of `bwt`?)

=======


It means that if we **standardize** `bwt`, the points become approximately normal with mean 0 and sd 1:

```{r fig.width=6, fig.height=3, dpi=150, cache=TRUE, fig.align='center'}
bwt.standardized = with(birthwt, (bwt - mean(bwt))/sd(bwt))

ggplot(mapping=aes(sample = bwt.standardized)) + stat_qq(distribution='qnorm') + 
  geom_abline(intercept=0, slope=1) + 
  labs(x = 'Normal Quantiles', y ='Birthweight (standardized)', title = 'QQ plot, Birth Weight (standardized) vs Normal Distribution')
```



Getting used to reading QQ plots
=======

QQ plots can take a lot of judgement to read -- how far from the x=y line is significant? 

Here are some new ideas that can help, from http://arxiv.org/abs/1503.02098 


What is this QQ plot saying?
=======

```{r fig.width=6, fig.height=3, dpi=150, cache=TRUE, fig.align='center', echo=FALSE}
# get the ratings for 2003
library(ggupset)
ratings.2003 = with(tidy_movies, rating[ year == 2003])
# standardize them
ratings.2003 = (ratings.2003 - mean(ratings.2003)) / sd(ratings.2003) 
```

Comparison of standardized 2003 ratings to a normal(0,1) distribution:

```{r fig.width=6, fig.height=3, dpi=150, cache=TRUE, fig.align='center', echo=FALSE}
# QQ plot vs normal dist
ggplot(mapping=aes(sample = ratings.2003)) + stat_qq(dist='qnorm') + geom_abline() + labs(title='QQ plot, ratings.2003 (standardized) \n vs normal distribution', x = 'normal quantiles', y = 'ratings quantiles')
```

The points are different from the line `y=x`, but how different? Is the difference big, or small?

One way of asking the question: could we have gotten a plot like this from normally distributed data? 

If so, then the difference is small.

=======

Here is what you get when you simulate data (of equal size) from a Normal(0,1) distribution and make a QQ plot.

```{r fig.width=6, fig.height=3, dpi=150, cache=TRUE, fig.align='center', echo=FALSE}
# generate normally distributed data
normal.data = rnorm(2158) # same length as ratings.2003
# QQ plot vs normal dist
ggplot(mapping = aes(sample = normal.data)) + stat_qq(dist = qnorm) + geom_abline() + labs(title = 'QQ plot, simulated data vs normal dist', x = 'normal quantiles', y = 'simulated data quantiles')
```

This looks pretty different from the previous plot

=======

v1 - v5 are all simulated data from a normal distribution. 

```{r fig.width=6, fig.height=4, dpi=150, cache=TRUE, fig.align='center', echo=FALSE}
# generate 5 sets of normal data, plus the 2003 ratings
#QQ.data = data.frame(v1 = rnorm(2158), v2 = rnorm(2158), v3 = rnorm(2158), v4 = rnorm(2158), v5 = rnorm(2158), ratings.2003 = ratings.2003)  
QQ.data = data.frame(v1 = rnorm(1920), v2 = rnorm(1920), v3 = rnorm(1920), v4 = rnorm(1920), v5 = rnorm(1920), ratings.2003 = ratings.2003)
# reshape for ggplot
QQ.data = melt(QQ.data)
# QQplot vs normal dist of each
ggplot(data = QQ.data, mapping=aes(sample = value)) + stat_qq(dist=qnorm) + facet_wrap('variable') + geom_abline()
```

If you can tell the difference, then `ratings.2003` is not the same distribution as v1 - v5

Are the birthweights normally distibuted?
=========

If we standardize the birthweights, do they become normally distributed?

```{r fig.width=6, fig.height=3, dpi=150, cache=TRUE, fig.align='center'}
# relabel birthwt$smoke so that it is more readable
birthwt$smoke = mapvalues(birthwt$smoke, from = c(0,1), to = c("no", "yes"))
# standardize the birth weights
birthwt = mutate(birthwt, bwt.std = (bwt - mean(bwt, na.rm=TRUE)) / sd(bwt, na.rm = TRUE))
```


========

One of these is `birthwt$bwt`. The rest are simulated Normal(0,1)

```{r fig.width=6, fig.height=5, dpi=150, cache=TRUE, fig.align='center', echo=FALSE}
# function: normal.df
# this function returns a data frame with a vector of normal data
normal.df = function(n.pts){
  return(data.frame(value = rnorm(n.pts)))  
}
# function Random.QQnorm
# lets you make a 8 QQ plots comparing simulated normal to a normal distribution, and 1 QQ plot comparing data.vec to a normal distribution (order is random)
Random.QQnorm = function(data.vec) {
  # how long is data.vec
  n.pts = length(data.vec)
  # make 9 sets of normal data of that length
  QQ.data = rdply(9, normal.df(n.pts))
  # which one do we replace with data.vec?
  random.choice = sample(1:9, size=1)
  # replace that one with data.vec
  QQ.data$value[ QQ.data$.n == random.choice ] = data.vec
  # return the data frame and the identity of data.vec
  return(list(data = QQ.data, correct = random.choice))
}
# generate 9 QQ plots, 1 of which is bwt.std
QQ.out = Random.QQnorm(birthwt$bwt.std)
# draw the plots
ggplot(data = QQ.out$data, mapping=aes(sample = value)) + stat_qq(dist=qnorm) + facet_wrap('.n', nrow = 3) + geom_abline() + labs(title = 'QQ plots', x = 'Normal Quantiles', y = 'Data Quantiles')
```

====

The correct answer was `r QQ.out$correct`.

Could you tell visually? 

If not, this suggests that `birthwt$bwt` cannot be distinguished from simulated data from a Normal(0,1) distribution

I.e., we fail to reject the hypothesis that `birthwt$bwt` is normal distributed.

```{r fig.width=6, fig.height=3, dpi=150, cache=TRUE, fig.align='center', echo=FALSE}
ggplot(mapping=aes(sample = birthwt$bwt.std)) + stat_qq(dist=qnorm) + geom_abline() + labs(title='QQ plot, standardized birthweights', x = 'Normal Quantiles', y = 'standardized birthweights')
```

Extra: Quantile Plots with `stat_qq()`
======

* `qunif()`: quantiles of a uniform distribution

* A QQ plot comparing a variable to the quantiles of a uniform distribution is the same thing as a quantile plot of the variable.

* that's why we can create a quantile plot using `stat_qq(distribution=qunif)`

* note: in older versions of R using `dist =` was the same as `distribution =`, but it seems to have been changed lately.


