---
title: Exercises Day 3
author: YOUR NAME HERE
output: html_document
---

**Note:** We will assume that the data file `expenditures_lec2.csv` is in your working directory.

Problem 1: Load the CSV file `expenditures_lec2.csv` into memory, using the command `read.file()` or `read.csv()` as described in lecture. Save the result as a variable.

```{r}
# Fill in, with comments added:

```

Problem 2: Create a scatter plot showing how `healthcare` costs behave as a function of the `age.respondent` variable. Don't forget to load any libraries you need!

```{r}
# Fill in, with comments added:

```

Problem 3: Create a bar chart showing how many rows in the data set are in each `region`. 

```{r'}
# Fill in, with comments added:

```

Problem 4: Use functions from lecture 2 (such as the `plyr` package) to compute the fraction (or percentage) of households that are in each `region`, instead of the raw counts. Then plot this as a bar chart.

```{r}
# Fill in, with comments added:

```

Problem 5: Remake the plot of problem 2, showing `healthcare` costs, but this time color-code the points to distinguish between those which spent over and under $3000 on health care.

```{r fig.width=6, fig.height=3, dpi=100, fig.align='center'}
# Fill in, with comments added:
```
